---
title: "Getting started with BARCS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with BARCS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r}
library(BARCS)
```

BARCS models a pooled CRISPR screen one guide at a time. For guide $i$ in
library $j$ it takes the guide's read count $K_{ij}$ against that library's
total mapped reads $n_j$, and fits

$$\operatorname{logit}(\mu_{ij}) = x_j'\beta_i, \qquad
\operatorname{Var}(K_{ij}) = n_j\mu_{ij}(1-\mu_{ij})\{1 + (n_j-1)\rho_i\}.$$

Two things follow from that variance, and they are the whole reason the
package exists.

First, the design is an arbitrary full-rank model matrix, not a pair of group
labels. Dose, time, an ordered phenotype, a batch or donor adjustment, and
interactions are all just columns of $x_j$.

Second, the beta-binomial term $\rho_i$ separates two kinds of noise that a
binomial model conflates: the sampling precision of sequencing, which grows
with depth, and the real variation between independently prepared libraries,
which does not. The reference distribution for a coefficient is a Student $t$
whose degrees of freedom come from the number of libraries minus the rank of
the design. **Sequencing more deeply never buys degrees of freedom.**

## The three objects

Every analysis carries the same three objects, and keeping them aligned is
essentially the whole discipline of using this package.

| Object | Shape | Rule |
|---|---|---|
| `counts` | guides × libraries | Raw integer read counts. |
| `totals` | one per library | Unfiltered mapped-guide totals. **Never recompute after filtering guides.** |
| `data` | one row per library | Same order as the columns of `counts`. |

The `totals` rule is the one that bites. It is the denominator of the model,
so it has to describe the library that was actually sequenced. If you drop
low-abundance guides from `counts` and then let `totals` default to the new
column sums, you have quietly changed the question every remaining guide is
answering.

## A worked screen

`evers_rt112` is a CRISPRn dropout screen: 961 guides across 93 genes,
sequenced before and after selection, with three libraries per group.

```{r}
data(evers_rt112)

str(evers_rt112$counts)
evers_rt112$design
```

Three libraries per group and two coefficients leaves four residual degrees of
freedom. That is a realistic amount of replication for a published screen, and
a useful reminder of the scale we are working at: the counts run to tens of
thousands of reads per guide, and none of that depth enters the $t$ reference.

```{r}
totals <- colSums(evers_rt112$counts)
totals
```

`bb_screen()` fits every guide against the shared design and reports one named
model-matrix coefficient. Because `group` has `"before"` as its reference
level, the coefficient `groupafter` reads as the log-odds change from before
to after selection.

```{r}
screen <- bb_screen(
  counts  = evers_rt112$counts,
  totals  = totals,
  data    = evers_rt112$design,
  formula = ~ group,
  term    = "groupafter",
  guide   = evers_rt112$guides$guide,
  gene    = evers_rt112$guides$gene
)

nrow(screen)
unique(screen$df)
head(screen[order(screen$p_value), c("gene", "estimate", "std_error", "p_value", "fdr")])
```

The most strongly depleted guides target proteasome and ribosome subunits,
which is what a dropout screen in a proliferating cell line should find.

## Read the controls before you read the hits

This screen ships with the reference gene sets the original study used: 46
genes expected to deplete, and 47 expected not to move. The second set is the
more useful one, because guides that should not move tell you what "not
moving" looks like in your data.

```{r}
control <- screen$gene %in% evers_rt112$nonessential

summary(screen$t_value[control])
mean(screen$p_value[control] < 0.05)
```

Almost every non-essential guide is "significant", and they lean positive.
That is not a bug in the model, and calibration is not the first thing to
reach for. It is a real feature of the experiment:

```{r}
round(totals / 1e6, 2)
```

The libraries shrink as selection kills cells, and the essential guides drop
out of them. A guide's abundance is measured *relative to its own library
total*, so every guide that merely survives gains share. Compositional shift
is intrinsic to the design, not noise.

What the negative controls give you is the size of that shift, expressed on
the same scale as the test statistics. `bb_calibrate_controls()` estimates one
scale factor from them and applies it to the whole screen.

```{r}
calibrated <- bb_calibrate_controls(screen, control = control, method = "qq_slope")

attr(calibrated, "control_scale")
mean(calibrated$p_value[control] < 0.05)
```

The control rejection rate drops from 95% to about 3% — the right order of
magnitude, and on the conservative side of the nominal 5%, which is the side
the `min_scale = 1` floor is designed to err on. The original columns are kept
with a `raw_` prefix, so nothing is lost:

```{r}
head(calibrated[, c("gene", "t_value", "raw_t_value", "p_value", "raw_p_value")], 3)
```

Two honest caveats. The `"qq_slope"` estimator used above fits the slope of
the control quantile-quantile plot over the 0.50–0.95 band; the default
`"tail_quantile"` matches a single far-tail order statistic and is noticeably
more variable with only a few hundred controls. And calibration only
*rescales* — it does not recentre. It works here because the shift is
absorbed into the scale, but a screen with a large offset and a small spread
would need the shift modelled instead, for instance with a normalisation term
in the design.

## Moderating the guide dispersion

With four residual degrees of freedom, each guide's own $\rho_i$ is a very
noisy estimate. `bb_moderate_dispersion()` shrinks each guide's variance
inflation toward a trend fitted across the whole library, using the scaled-F
moment estimator of Smyth (2004), and returns a moderated $t$ on
`df_residual + prior_df` degrees of freedom.

```{r}
moderated <- bb_moderate_dispersion(screen)

attr(moderated, "prior_df")
attr(moderated, "df_total")
```

The prior is worth many times the four degrees of freedom the design alone
supports, which is exactly why moderation is worth doing at this replication
level. It is independent of control calibration; the two address different
problems and can be combined.

## Beyond two groups

Nothing above needed the design to be two groups. Here is a dose–response
screen with a batch covariate, which the original two-group beta-binomial test
could not express at all.

```{r}
set.seed(20260801)

design <- data.frame(
  sample_name = sprintf("sample_%02d", 1:18),
  dose  = rep(seq(-1.5, 1.5, length.out = 6), 3),
  batch = factor(rep(letters[1:3], each = 6))
)
lib_totals <- round(runif(nrow(design), 40000, 90000))

simulate_guide <- function(intercept, slope, rho = 0.0015) {
  mu <- plogis(intercept + slope * design$dose)
  precision <- 1 / rho - 1
  p <- rbeta(length(mu), mu * precision, (1 - mu) * precision)
  rbinom(length(mu), lib_totals, p)
}

counts <- rbind(
  geneA_sg1 = simulate_guide(-7.0, 0.9),
  geneA_sg2 = simulate_guide(-6.8, 0.8),
  geneB_sg1 = simulate_guide(-7.1, 0.0),
  geneB_sg2 = simulate_guide(-6.9, 0.0)
)
```

One guide at a time, with `summary()` and a contrast:

```{r}
fit <- bbreg(
  count   = counts[1, ],
  total   = lib_totals,
  formula = ~ dose + batch,
  data    = design
)

summary(fit)
```

`bb_contrast()` tests any linear combination of the fitted coefficients. Here
is the difference between batches `b` and `c`, which no single coefficient
reports directly:

```{r}
bb_contrast(fit, c(batchb = -1, batchc = 1))
```

And the whole screen at once:

```{r}
bb_screen(
  counts  = counts,
  totals  = lib_totals,
  data    = design,
  formula = ~ dose + batch,
  term    = "dose",
  gene    = sub("_sg[0-9]+$", "", rownames(counts))
)
```

The `dose` coefficient is a log-odds change per unit of dose. Its $t$
reference has 18 − 4 = 14 residual degrees of freedom, from the eighteen
libraries, not from the tens of thousands of reads in each.

## Where to go next

* `vignette("barcs-quantification")` — turning FASTQ files into the `counts`
  and `totals` this vignette started from, and the library QC worth doing
  before any modelling.
* `vignette("barcs-gene-summaries")` — the `bb_gene_*()` functions, what null
  each one assumes, and why guide count is not biological replication.

## References

Baggerly KA, Deng L, Morris JS, Aldaz CM (2004). Overdispersed logistic
regression for SAGE. *BMC Bioinformatics*, 5, 144.

Evers B, Jastrzebski K, Heijmans JPM, Grernrum W, Beijersbergen RL, Bernards R
(2016). CRISPR knockout screening outperforms shRNA and CRISPRi in identifying
essential genes. *Nature Biotechnology*, 34(6), 631–633.

Jeong H-H, Kim SY, Rousseaux MWC, Zoghbi HY, Liu Z (2019). Beta-binomial
modeling of CRISPR pooled screen data identifies target genes with greater
sensitivity and fewer false negatives. *Genome Research*, 29(6), 999–1008.

Smyth GK (2004). Linear models and empirical Bayes methods for assessing
differential expression in microarray experiments. *Statistical Applications
in Genetics and Molecular Biology*, 3(1), 3.
