---
title: "Summarising guides to genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising guides to genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r}
library(BARCS)
```

BARCS fits guides, but screens are usually read as gene lists. Getting from
one to the other requires an assumption, and the assumption is where most of
the statistical trouble in this field lives.

## The thing that is not true

The tempting move is to treat a gene's guides as replicates: five guides, five
observations, four degrees of freedom, done. They are not replicates. Guides
targeting the same gene share every library, every cell passage, and every
sequencing run. What varies between them is guide efficiency and off-target
activity — real biology, but not the biological variation that a p-value about
the gene is supposed to be about.

Treating them as replicates produces confident, reproducible, wrong answers,
and it gets worse as you add guides. BARCS therefore does not offer a single
`gene_test()`. It offers five summaries, each named for what it actually
assumes, and each stamping its assumption onto the object it returns.

| Function | Combines | Assumes |
|---|---|---|
| `bb_gene_original()` | Signed $z$ from guide p-values | The historical BARCS statistic; kept for continuity |
| `bb_gene_normal()` | Guide coefficients, equally weighted | Guide effects are exchangeable draws from $N(\mu_g,\sigma_g^2)$ |
| `bb_gene_consistency()` | Inverse-variance weighted effect | One shared gene effect; calibrated against an empirical null |
| `bb_gene_partial_pool()` | Random-effects weights | Guide heterogeneity $\tau_g^2$ estimated per gene (DerSimonian–Laird) |
| `bb_gene_eb_moderate()` | Random-effects, shrunken $\tau_g^2$ | As above, with $\tau_g^2$ shrunk toward a screen-wide prior |

None of them combines guide p-values by Fisher's or Stouffer's method across
guides treated as independent experiments.

## A quick tour

`barcs_example_guides()` returns a fixed table in the shape `bb_screen()`
produces: two all-control genes, one gene whose guides agree, and one whose
guides disagree.

```{r}
guides <- barcs_example_guides()
control <- startsWith(guides$gene, "control")

guides[guides$gene %in% c("consistent", "discordant"), c("gene", "estimate", "std_error")]
```

Both genes have the same mean effect if you only look at the concordant ones,
but `discordant` has guides pointing in opposite directions. A method that
ignores heterogeneity will rank them similarly; one that models it will not.

```{r}
pooled <- bb_gene_partial_pool(guides, control = control, min_control_genes = 2)
pooled[, c("gene", "estimate", "tau2", "i_squared", "guide_direction_agreement", "p_value")]
```

`tau2` is the estimated between-guide variance and `i_squared` the fraction of
total variance it accounts for. The discordant gene absorbs a large `tau2`,
its guides get down-weighted, and its p-value reflects the disagreement.

Each result also carries diagnostics for whether one guide is driving the
answer:

```{r}
pooled[, c("gene", "max_weight_fraction", "leave_one_out_max_change", "leave_one_out_sign_stable")]
```

A gene whose `max_weight_fraction` is near one is a gene with a single
informative guide, whatever its p-value says.

## On a real screen

```{r}
data(evers_rt112)

screen <- bb_screen(
  counts  = evers_rt112$counts,
  totals  = colSums(evers_rt112$counts),
  data    = evers_rt112$design,
  formula = ~ group,
  term    = "groupafter",
  guide   = evers_rt112$guides$guide,
  gene    = evers_rt112$guides$gene
)

is_control <- screen$gene %in% evers_rt112$nonessential
```

This screen has reference answers: 46 genes that should deplete and 47 that
should not. That lets us separate two questions that are usually confused —
whether a method *ranks* well, and whether its *FDR* means anything.

```{r}
essential <- evers_rt112$essential

auc <- function(result) {
  score <- -result$statistic
  label <- result$gene %in% essential
  ranks <- rank(score)
  n1 <- sum(label); n0 <- sum(!label)
  (sum(ranks[label]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

summarise <- function(result, name) {
  hits <- result$gene[result$fdr < 0.05]
  data.frame(
    method       = name,
    genes        = nrow(result),
    auc          = auc(result),
    hits         = length(hits),
    hits_true    = sum(hits %in% essential),
    hits_false   = sum(hits %in% evers_rt112$nonessential)
  )
}

comparison <- rbind(
  summarise(bb_gene_original(screen), "original"),
  summarise(bb_gene_normal(screen), "normal"),
  summarise(bb_gene_consistency(screen, control = is_control), "consistency"),
  summarise(bb_gene_partial_pool(screen, control = is_control), "partial_pool"),
  summarise(bb_gene_eb_moderate(screen, control = is_control), "eb_moderate")
)
comparison
```

**Every method ranks perfectly.** An AUC of 1 means every essential gene
ranks above every non-essential gene, whichever summary you use. For the usual
purpose of a screen — deciding what to follow up — the choice barely matters
here. (`normal` and `consistency` report 92 genes rather than 93 because they
default to requiring three usable guides per gene.)

**No method's FDR is trustworthy on this dataset,** and the two families fail
in opposite directions. `original` and `normal` call nearly everything,
including all 47 genes that should not move. The three empirical-null methods
call nothing.

Both failures have the same root cause, and it is worth understanding rather
than tuning away.

```{r}
summary(screen$t_value[is_control])
```

The non-essential control guides are centred around $t \approx 9$, not zero.
As `vignette("BARCS")` explains, that is compositional: the libraries shrink
during selection, so guides that merely survive gain share. `original` and
`normal` take those inflated guide statistics at face value and duly declare
the controls significant.

The empirical-null methods notice the problem and correct for it, but
overcorrect:

```{r}
pool <- bb_gene_partial_pool(screen, control = is_control)
c(
  null_center   = attr(pool, "null_center"),
  global_scale  = attr(pool, "global_scale"),
  control_scale = attr(pool, "control_scale"),
  null_scale    = attr(pool, "null_scale")
)
```

The scale actually used is the largest of the candidates, and here that is
`global_scale`, the median absolute deviation across all genes. That estimator
assumes fewer than half the genes are active. This is a *benchmark* library,
deliberately built as roughly half essential and half not, so the assumption
fails and the all-gene MAD absorbs the real signal into the null.

The lesson is not that one method wins. It is that a benchmark library and a
genome-wide library are different objects, and the robust-null machinery is
built for the second. On a real library, where the overwhelming majority of
genes do nothing, `global_scale` is a reasonable null estimate and this
conservatism does not appear.

## Practical guidance

* **Start with the ranking.** It is the most robust output BARCS produces and
  the least sensitive to which summary you pick.
* **Check the controls before the hits.** If the non-essential guides are not
  centred near zero, no gene-level FDR downstream will mean anything until you
  understand why.
* **Prefer `bb_gene_eb_moderate()` for screens with few guides per gene.** Per
  gene, $\tau_g^2$ is estimated from four or five numbers; shrinking it toward
  a screen-wide prior is strictly better conditioned.
* **Use `bb_gene_partial_pool()` when guides per gene are plentiful** and you
  would rather each gene's heterogeneity speak for itself.
* **Read the diagnostics.** `guide_direction_agreement`,
  `max_weight_fraction`, and `leave_one_out_sign_stable` catch the
  single-guide artefacts that a p-value hides.
* **Reach for `bb_gene_original()` only for continuity** with earlier BARCS
  and CB2 analyses.

## A note on double correction

The pooling functions deliberately prefer the `raw_std_error` column when
`bb_calibrate_controls()` has already run:

```{r}
calibrated <- bb_calibrate_controls(screen, control = is_control, method = "qq_slope")

identical(
  bb_gene_partial_pool(calibrated, control = is_control)$raw_statistic,
  bb_gene_partial_pool(screen, control = is_control)$raw_statistic
)
```

This is intentional. These methods fit their own gene-level empirical null, so
consuming already-widened standard errors would penalise the same inflation
twice. `bb_gene_original()` is the exception — it reads the p-value column
directly, so it does respond to guide-level calibration, which is what you
want from a method that has no null model of its own.

## References

DerSimonian R, Laird N (1986). Meta-analysis in clinical trials. *Controlled
Clinical Trials*, 7(3), 177–188.

Smyth GK (2004). Linear models and empirical Bayes methods for assessing
differential expression in microarray experiments. *Statistical Applications
in Genetics and Molecular Biology*, 3(1), 3.
