---
title: "From FASTQ files to a count matrix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From FASTQ files to a count matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r}
library(BARCS)
```

`barcs_quant()` turns raw sequencing reads into the guide-by-library count
matrix that `bb_screen()` consumes. It needs two things: a FASTA file
describing the guide library, and a data frame saying which FASTQ file belongs
to which sample.

## How matching works

Each guide sequence is packed into a 2-bit-per-base integer and stored in a
hash table. Every read is then scanned with a rolling window the width of a
guide, and the first window that hits the table wins.

The practical consequence is that **you do not have to tell BARCS where the
guide sits inside the read**. There is no adapter argument, no constant-region
offset, and no trimming step. A read with a variable-length stagger, or with
the vector sequence still attached, matches just as well as a trimmed one.

Matching is exact — no mismatches are allowed. For 20-mers at typical
sequencing quality this loses few reads, and it means a count can never be
credited to the wrong guide because of a single base error.

## A worked example

The package ships a small synthetic screen: six FASTQ files and a 25-guide
library.

```{r}
fasta   <- system.file("extdata", "toydata", "small_sample.fasta", package = "BARCS")
toydata <- system.file("extdata", "toydata", package = "BARCS")

samples <- c("Base1", "Base2", "High1", "High2", "Low1", "Low2")
design <- data.frame(
  sample_name = samples,
  group       = rep(c("Base", "High", "Low"), each = 2),
  fastq_path  = file.path(toydata, paste0(samples, ".fastq.gz"))
)
design
```

Only `sample_name` and `fastq_path` are required; any other columns are
carried through and are available later as model covariates. Files ending in
`.gz` are decompressed to a temporary file automatically.

```{r}
quantified <- barcs_quant(fasta, design)
quantified
```

```{r}
head(quantified$counts)
```

## `totals` is not `reads`

This distinction matters enough to be worth its own section, because getting
it wrong silently changes every p-value in the screen.

```{r}
rbind(
  totals = quantified$totals,
  reads  = quantified$reads
)
```

* `reads` is the number of FASTQ records in the file. It includes everything
  that failed to match: adapter dimers, contamination, reads off the end of
  the library.
* `totals` is the number of reads that carried a recognisable guide barcode.
  It equals the column sums of the unfiltered count matrix.

`totals` is the beta-binomial denominator, and it is what you pass to
`bb_screen()`. `reads` exists so you can compute mappability. Feeding `reads`
to the model instead would treat unmappable sequencing as though it were
library that a guide could have occupied.

```{r}
barcs_mappability(quantified)
```

These are synthetic reads, so they all map. On real data the number to watch
is consistency: one sample far below its peers usually means the wrong library
FASTA, a trimming problem upstream, or contamination, and is worth resolving
before any modelling.

## Library QC

`barcs_quant()` reports what it had to discard from the library, rather than
quietly working around it.

```{r}
str(quantified$library)
```

Three classes of entry are dropped:

* **Repeated sequences.** If two guides share a sequence, a read matching it
  cannot be attributed to either, so both entries go. The package includes a
  FASTA with a deliberate duplicate to show this:

```{r}
duplicated_library <- barcs_quant(
  system.file("extdata", "toydata", "small_sample_dup.fasta", package = "BARCS"),
  design[1, ]
)
duplicated_library$library$n_duplicate
```

* **Off-length guides.** The scanner uses one window width, taken as the
  library's modal guide length, so entries of any other length are reported
  and skipped.
* **Non-ACGT sequences.** Degenerate IUPAC codes are refused rather than
  coerced to a base.

A library where these counts are not what you expect is a library worth
looking at before trusting the counts.

## Strand orientation

Libraries are sequenced in either orientation depending on the primer design.
Reads that fail to match are automatically rescanned as their reverse
complement, and whichever orientation matched more reads decides the reported
counts.

```{r}
quantified$reverse_complement
```

You do not need to know the orientation in advance, and you do not need to
pre-process the FASTQ. If this flag comes back `TRUE` for some samples and
`FALSE` for others in the same experiment, that is worth investigating — it
usually means the samples were prepared with different primers.

## Attaching gene annotation

Guide-to-gene mapping can come from either of two places. If your guide names
already encode the gene, derive it directly:

```{r}
gene <- sub("_[^_]+$", "", quantified$guides$guide)
head(data.frame(guide = quantified$guides$guide, gene = gene))
```

Otherwise pass a CSV or TSV to `map_path`. It needs a `gene` column and a
guide identifier column named `id`, `sgRNA`, or `guide`:

```{r}
annotated <- barcs_quant(
  fasta, design,
  map_path = file.path(toydata, "sg2gene.csv")
)
head(annotated$guides)
```

Guides with no entry in the mapping file get `NA` and a warning, rather than
being dropped.

## Handing off to the model

The quantification object holds everything `bb_screen()` needs, and the design
comes back exactly as supplied so the column order is guaranteed to line up.

```{r}
screen <- bb_screen(
  counts  = annotated$counts,
  totals  = annotated$totals,
  data    = annotated$design,
  formula = ~ group,
  term    = "groupHigh",
  gene    = annotated$guides$gene,
  min_total_count = 0
)
head(screen[, c("gene", "guide", "estimate", "std_error", "p_value")])
```

With two libraries per group and three groups this toy screen has three
residual degrees of freedom, so the p-values are illustrative only.

If you do filter guides before modelling — which is often sensible for guides
with almost no reads — keep the totals from the unfiltered matrix:

```{r}
keep <- rowSums(annotated$counts) >= 10
sum(keep)

# Correct: the denominator still describes the library that was sequenced.
filtered <- bb_screen(
  counts  = annotated$counts[keep, ],
  totals  = annotated$totals,
  data    = annotated$design,
  formula = ~ group,
  term    = "groupHigh",
  min_total_count = 0
)
nrow(filtered)
```

## Performance

Scanning is linear in the number of reads and effectively constant per read.
On Unix-like systems `ncores` forks workers across samples:

```{r, eval = FALSE}
quantified <- barcs_quant(fasta, design, ncores = 4)
```

Each worker builds its own copy of the guide index, so this trades a little
memory for close to linear speedup across samples. Windows always uses one
worker. Long scans check for user interrupts, so a run against a full-size
FASTQ can be stopped from the console.
