---
title: "Summarising guides to genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising guides to genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r}
library(BARCS)
```

BARCS fits guides, but screens are usually read as gene lists. Getting from
one to the other requires an assumption, and the assumption is where most of
the statistical trouble in this field lives.

## The thing that is not true

The tempting move is to treat a gene's guides as replicates: five guides, five
observations, four degrees of freedom, done. They are not replicates. Guides
targeting the same gene share every library, every cell passage, and every
sequencing run. What varies between them is guide efficiency and off-target
activity — real biology, but not the biological variation that a p-value about
the gene is supposed to be about.

Treating them as replicates produces confident, reproducible, wrong answers,
and it gets worse as you add guides. BARCS therefore does not offer a
`gene_test()` that pretends otherwise. Its recommended summary,
`bb_gene_stouffer()`, combines guide *tests*, not guide replicates, and says
plainly what its gene-level p-value assumes. Four effect-pooling alternatives
are kept as experimental sensitivity analyses.

| Function | Status | Combines | Assumes |
|---|---|---|---|
| `bb_gene_stouffer()` | **Recommended** | Signed $z$ from guide p-values; median guide effect | Null guide scores are independent |
| `bb_gene_normal()` | Experimental | Guide coefficients, equally weighted | Guide effects are exchangeable draws from $N(\mu_g,\sigma_g^2)$ |
| `bb_gene_consistency()` | Experimental | Inverse-variance weighted effect | One shared gene effect; calibrated against an empirical null |
| `bb_gene_partial_pool()` | Experimental | Random-effects weights | Guide heterogeneity $\tau_g^2$ estimated per gene (DerSimonian–Laird) |
| `bb_gene_eb_moderate()` | Experimental | Random-effects, shrunken $\tau_g^2$ | As above, with $\tau_g^2$ shrunk toward a screen-wide prior |

`bb_gene_stouffer()` is the default because it ranked genes best in simulated
FACS screens with known truth, ahead of all four alternatives on average
precision and F1 in most scenarios. Its weakness is calibration rather than
ranking: guides targeting one gene share every library, so correlated guide
scores make its $\sqrt{m_g}$ denominator too small and its gene-level p-values
anti-conservative. Rank with it freely; check its FDR against negative-control
genes aggregated the same way before reporting one.

## A quick tour

`barcs_example_guides()` returns a fixed table in the shape `bb_screen()`
produces: two all-control genes, one gene whose guides agree, and one whose
guides disagree.

```{r}
guides <- barcs_example_guides()
control <- startsWith(guides$gene, "control")

guides[guides$gene %in% c("consistent", "discordant"), c("gene", "estimate", "std_error")]
```

Both genes have the same mean effect if you only look at the concordant ones,
but `discordant` has guides pointing in opposite directions. A method that
ignores heterogeneity will rank them similarly; one that models it will not.

```{r}
pooled <- bb_gene_partial_pool(guides, control = control, min_control_genes = 2)
pooled[, c("gene", "estimate", "tau2", "i_squared", "guide_direction_agreement", "p_value")]
```

`tau2` is the estimated between-guide variance and `i_squared` the fraction of
total variance it accounts for. The discordant gene absorbs a large `tau2`,
its guides get down-weighted, and its p-value reflects the disagreement.

Each result also carries diagnostics for whether one guide is driving the
answer:

```{r}
pooled[, c("gene", "max_weight_fraction", "leave_one_out_max_change", "leave_one_out_sign_stable")]
```

A gene whose `max_weight_fraction` is near one is a gene with a single
informative guide, whatever its p-value says.

## On a real screen

```{r}
data(evers_rt112)

screen <- bb_screen(
  counts  = evers_rt112$counts,
  totals  = colSums(evers_rt112$counts),
  data    = evers_rt112$design,
  formula = ~ group,
  term    = "groupafter",
  guide   = evers_rt112$guides$guide,
  gene    = evers_rt112$guides$gene
)

is_control <- screen$gene %in% evers_rt112$nonessential
```

This screen has reference answers: 46 genes that should deplete and 47 that
should not. That lets us separate two questions that are usually confused —
whether a method *ranks* well, and whether its *FDR* means anything.

```{r}
essential <- evers_rt112$essential

auc <- function(result) {
  score <- -result$statistic
  label <- result$gene %in% essential
  ranks <- rank(score)
  n1 <- sum(label); n0 <- sum(!label)
  (sum(ranks[label]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

summarise <- function(result, name) {
  hits <- result$gene[result$fdr < 0.05]
  data.frame(
    method       = name,
    genes        = nrow(result),
    auc          = auc(result),
    hits         = length(hits),
    hits_true    = sum(hits %in% essential),
    hits_false   = sum(hits %in% evers_rt112$nonessential)
  )
}

comparison <- rbind(
  summarise(bb_gene_stouffer(screen), "stouffer"),
  summarise(bb_gene_normal(screen), "normal"),
  summarise(bb_gene_consistency(screen, control = is_control), "consistency"),
  summarise(bb_gene_partial_pool(screen, control = is_control), "partial_pool"),
  summarise(bb_gene_eb_moderate(screen, control = is_control), "eb_moderate")
)
comparison
```

**Every method ranks perfectly.** An AUC of 1 means every essential gene
ranks above every non-essential gene, whichever summary you use. For the usual
purpose of a screen — deciding what to follow up — the choice barely matters
here. (`normal` and `consistency` report 92 genes rather than 93 because they
default to requiring three usable guides per gene.)

**No method's FDR is trustworthy on this dataset,** and the two families fail
in opposite directions. `stouffer` and `normal` call nearly everything,
including all 47 genes that should not move. The three empirical-null methods
call nothing.

Both failures have the same root cause, and it is worth understanding rather
than tuning away.

```{r}
summary(screen$t_value[is_control])
```

The non-essential control guides are centred around $t \approx 9$, not zero.
As `vignette("BARCS")` explains, that is compositional: the libraries shrink
during selection, so guides that merely survive gain share. `stouffer` and
`normal` take those inflated guide statistics at face value and duly declare
the controls significant.

The empirical-null methods notice the problem and correct for it, but
overcorrect:

```{r}
pool <- bb_gene_partial_pool(screen, control = is_control)
c(
  null_center   = attr(pool, "null_center"),
  global_scale  = attr(pool, "global_scale"),
  control_scale = attr(pool, "control_scale"),
  null_scale    = attr(pool, "null_scale")
)
```

The scale actually used is the largest of the candidates, and here that is
`global_scale`, the median absolute deviation across all genes. That estimator
assumes fewer than half the genes are active. This is a *benchmark* library,
deliberately built as roughly half essential and half not, so the assumption
fails and the all-gene MAD absorbs the real signal into the null.

The lesson is not that one method wins. It is that a benchmark library and a
genome-wide library are different objects, and the robust-null machinery is
built for the second. On a real library, where the overwhelming majority of
genes do nothing, `global_scale` is a reasonable null estimate and this
conservatism does not appear.

The better fix is upstream of any gene summary: remove the composition shift
with control totals, as shown in `vignette("BARCS")`, and then calibrate the
guide statistics on controls before passing them to `bb_gene_stouffer()`.

## Practical guidance

* **Use `bb_gene_stouffer()`.** It is the recommended summary and the one
  benchmarked in the BARCS manuscript.
* **Start with the ranking.** It is the most robust output BARCS produces and
  the least sensitive to which summary you pick.
* **Check the controls before the hits.** If the non-essential guides are not
  centred near zero, no gene-level FDR downstream will mean anything until you
  understand why; consider `barcs_control_totals()`.
* **Check the gene-level FDR on control genes aggregated the same way.** Group
  negative-control guides into pseudo-genes with the same number of guides as
  real genes and confirm that few of them are called. Correlated guides inflate
  Stouffer gene p-values even when the guide-level tests are well calibrated.
* **Treat the four experimental summaries as sensitivity analyses.** Their
  diagnostics — `guide_direction_agreement`, `max_weight_fraction`, and
  `leave_one_out_sign_stable` — are still useful for spotting a gene driven by
  a single guide.

## A note on double correction

The pooling functions deliberately prefer the `raw_std_error` column when
`bb_calibrate_controls()` has already run:

```{r}
calibrated <- bb_calibrate_controls(screen, control = is_control, method = "qq_slope")

identical(
  bb_gene_partial_pool(calibrated, control = is_control)$raw_statistic,
  bb_gene_partial_pool(screen, control = is_control)$raw_statistic
)
```

This is intentional. These methods fit their own gene-level empirical null, so
consuming already-widened standard errors would penalise the same inflation
twice. `bb_gene_stouffer()` is the exception — it reads the p-value column
directly, so it does respond to guide-level calibration, which is what you
want from a method that has no null model of its own.

## References

DerSimonian R, Laird N (1986). Meta-analysis in clinical trials. *Controlled
Clinical Trials*, 7(3), 177–188.

Smyth GK (2004). Linear models and empirical Bayes methods for assessing
differential expression in microarray experiments. *Statistical Applications
in Genetics and Molecular Biology*, 3(1), 3.
